---
title: "01: Vectors, Matrices, and Linear Transformations"
author: "Anna Astakhova"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../R/matrix_ops.R")
```

# Motivation

In data science, a structured dataset is formally represented as a matrix where rows correspond to discrete observations and columns correspond to variables. Because of this inherent mathematical structure, every fundamental operation performed on data, whether scaling features, calculating covariance, or projecting high-dimensional datasets into lower-dimensional subspaces via Principal Component Analysis, is a matrix operation. Understanding linear algebra provides the rigorous framework necessary to understand the mechanics governing these computational algorithms.

# Vectors and the Dot Product

A vector is formally defined as an element of a vector space subject to the operations of vector addition and scalar multiplication. While sometimes visualized in two or three dimensions as a directed line segment, in data science it strictly represents a specific coordinate position in an n-dimensional space.

The dot product of two vectors $u$ and $v$ is:

$$ u*v = \sum_{i=1}^{n} u_i v_i = |u||v| cos\theta$$
where θ is the angle between the vectors in Euclidean space. When the dot product evaluates to exactly zero, the vectors are orthogonal, meaning they are linearly independent and span non-overlapping dimensions.



```{r}

u <- c(1, 2, 3)
v <- c(4, 5, 6)

# our implementation
dot_product(u, v)

# algebraic check: standard basis vectors should be orthogonal
dot_product(c(1, 0), c(0, 1))
```

The second result evaluates to exactly 0, confirming that the standard basis vectors are orthogonal.

# Matrix Multiplication

A matrix fundamentally represents a linear transformation (or linear map) between vector spaces. Matrix multiplication represents the mathematical composition of these linear transformations.

When calculating the product of matrix $A$ and matrix $B$, the entry in the $i$-th row and $j$-th column of the resulting matrix is computed as the dot product of the $i$-th row of $A$ and the $j$-th column of $B$.
This algebraically necessitates the rule of inner dimensions: the operation is only defined if the number of columns in $A$ exactly equals the number of rows in $B$.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2, byrow = TRUE)
B <- matrix(c(5, 6, 7, 8), nrow = 2, byrow = TRUE)

# Calculate using our scratch implementation
mat_mul(A, B)

# Verify against R's built-in operator
A %*% B
```

Because matrix multiplication defines the composition of two mappings, the operation is fundamentally non-commutative ($AB \neq BA$). Applying transformation $A$ followed by $B$ generally yields a different mathematical mapping than applying $B$ followed by $A$.

# Determinants

The determinant is a scalar property of a square matrix that encodes fundamental information about the linear transformation it represents. Geometrically, it quantifies the n-dimensional volume scaling factor of that transformation.

A determinant of exactly 0 indicates a singular matrix. This means the matrix's column vectors are linearly dependent, and the transformation maps the original space onto a lower-dimensional subspace (indicating a non-trivial null space). Because dimensionality is lost, the mapping is not bijective and no inverse transformation can exist.

For a 2x2 matrix, the determinant is defined algebraically as:

$$ det(A) = ad - bc$$
For a 3x3 matrix, it is computed via Laplace expansion (cofactor expansion) along the first column, which recursively reduces the operation to smaller 2x2 sub-determinants:

$$ det(A) = a(M_{11}) - b(M_{21}) - c(M_{31}) $$
```{r}
M2 <- matrix(c(4, 2, 1, 3), nrow = 2, byrow = TRUE)
M3 <- matrix(c(1, 2, 3, 0, 1, 4, 5, 6, 0), nrow = 3, byrow = TRUE)

# 2x2 Determinant
det_2x2(M2)
det(M2) # R's implementation

# 3x3 Determinant
det_3x3(M3)
det(M3) # R's implementation
```

# The Inverse

The inverse of a square matrix $A$, denoted $A^{-1}$, is the unique matrix such that their product yields the identity matrix $I$. Algebraically, this is expressed as $AA^{-1}=I$. It represents the linear transformation that precisely reverses the mapping applied by $A$.

For a 2x2 matrix, the inverse is derived via the closed-form formula:

$$ A^{-1} = \frac{1}{det(A)} * \begin{bmatrix} d & -b \\ -c & a \end{bmatrix} $$
```{r}
# Calculate the inverse of M2 using our implementation
M2_inv <- inv_2x2(M2)
M2_inv

# Verify that A %*% A_inv equals the identity matrix
is_inverse(M2, M2_inv)
```

As demonstrated by the scalar multiplier $\frac{1}{det(A)}$ in the formula, invertibility is strictly dependent on the determinant. If $det(A) = 0$, the matrix is singular and the division is mathematically undefined, confirming that lower-dimensional projections cannot be uniquely reversed.

# What We Built

In this vignette we implemented vectors, dot products, matrix multiplication, determinants, and inverses from scratch without using R's built-in functions, and validated each against R's standard implementations. In the next vignette we use these foundations to solve systems of linear equations.
